Two-Phase Commit Using Blockchain

### 2022-06-04 paper/paper.tex
Update draft (citation hasn't completed yet).

\documentclass[11pt,conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

We describe the details of each component in the remaining \ref{components}, the complete flow of handling a successful transaction in \ref{tnxflow}, and how we handle failures in \ref{failurecases}.

% \begin{figure}[h]
  % \centering
  % \includegraphics[width=0.4\textwidth]{./diagram.png}
% \caption{Fig 1. This diagram displays the main components within our architecture and example steps of successfully committing a transaction.}
  % \label{fig:Panel}
% \end{figure}

@@

\subsubsection{Coordinator} \label{coordinator}
The coordinator exposes two APIs to the clients:
\begin{itemize}
  \item \texttt{CommitAtomicTransaction(CommitAtomicTransactionRequest, CommitAtomicTransactionResponse)}
  \item \texttt{GetTransactionResult(GetTransactionResultRequest, GetTransactionResultResponse)}
\end{itemize}

@@

\subsubsection{Cohort} \label{cohort}
The cohort exposes two APIs to the coordinator:
\begin{itemize}
  \item \texttt{PrepareTransaction(
    PrepareTransactionRequest, PrepareTransactionResponse)}
  \item \texttt{GetTransactionResult(
    GetTransactionResultRequest, GetTransactionResultResponse)}
\end{itemize}

The $PrepareTransactionRequest$ contains a $transaction\_id$, a list of operations of the sub-transaction and a vote timeout time, after which the blockchain will not accept votes. When a cohort receives a \texttt{PrepareTransaction} request from the coordinator, it takes these steps: 
\begin{enumerate}
  \item Start a new thread in the thread pool.
  \item Create a \texttt{DatabaseAdapter} object.
  \item Start a transaction with 
    \texttt{DatabaseAdapter\\::Begin()}. 
  \item Acquire the relevant locks for the sub-transaction. 
  \item Execute a series of requested Put or Get operations in the sub-transaction. 
    If any operation is reported invalid by the database or locks cannot be acquired in time, the cohort will send an $ABORT$ ballot to the blockchain via \texttt{BlockchainAdapter::Vote()}.
  \item If all operations succeed, it will store the retrieved Get values in an in-memory hashmap as well as persistently on disk keyed by the $transaction\_id$ to prepare for future \texttt{GetTransactionResult} calls.
  \item Set a timer based on the vote timeout time.
  \item If the timer reaches the vote timeout time, it will send a request \texttt{BlockchainAdapter::\\GetVotingDecision(transaction\_id)} to the blockchain. 
    \begin{enumerate}
      \item If the blockchain states that the transaction should be committed, it will call \texttt{DatabaseAdapter::Commit()}.
      \item If the blockchain states that the transaction should be aborted, it will call \texttt{DatabaseAdapter::Abort()}.
    \end{enumerate}

\end{enumerate}

When the cohort receives a \texttt{GetTransaction\-Result} request and has committed the transaction, it simply reads the in-memory hashmap based on the $transaction\_id$ in the request and returns the got values as response. If it has not yet received a commit or abort response from the blockchain, it returns a pending response instead.

\subsection{System Interactions (a.k.a Life of a Transaction)} \label{txnflow}
This section describes the end-to-end life of a transaction without handling failures.

\begin{enumerate}
  \item The client composes a transaction request and passes it to the coordinator through \texttt{\\CommitAomticTransaction}. The coordinator returns a transaction ID for the client to look up the transaction result later.
  \item The coordinator asks the blockchain to trace votes for the new transaction through \texttt{StartVoting}.
  \item The coordinator decomposes the transaction into sub-transactions and distributes them to the corresponding cohort through \texttt{PrepareTransaction}.
  \item Each cohort executes a sub-transaction by acquiring locks and forwarding the operations to the database up to the point before Commit(). The cohort writes an entry to the redo log to recover in case it crashes. Additionally, the cohort writes the \texttt{GET} operation results to disk for faster responses after a crash and recovery.
  \item The cohort votes to the blockchain through \texttt{Vote} indicating that it wants to commit or abort the transaction.
  \item The cohort waits until the expected transaction timeout time and gets the voting decision from the blockchain through \texttt{GetVotingDecision}.
  \item If the voting decision is to commit the transaction, cohort commits the sub-transaction onto the database and releases any acquired locks; otherwise, it aborts the sub-transaction in the database and releases any acquired locks.
  \item (Independent of the previous steps) The client waits until the timeout time and requests the transaction result from the coordinator through \texttt{GetTransactionResult}, using the transaction ID given in step 1.
  \item The coordinator gets (1) the voting decision of the inquiry transaction from the blockchain through \texttt{GetTransactionResult} and (2) the GET operation results from each cohort.
\end{enumerate}

@@

\subsubsection{Crash of the coordinator}
The state of the coordinator includes only the list of which cohorts are involved in each active transaction. Thus if the coordinator crashes, when it recovers, it will respond to client requests for the \texttt{GET} responses of the transaction by indicating that it is unable to find the transaction. As a result, the client can directly ask the cohorts what the \texttt{GET} responses of the transaction are. Unlike in traditional 2PC, where the coordinator stores how each cohort has voted, in our design, this information is stored in the blockchain. Our system is resilient to such crashes, as the transaction does not need to block on the coordinator’s recovery.

\subsubsection{Crash of a cohort}
Our system is also resilient to crashes of cohorts, as these crashes do not cause any other cohorts to block for longer than the vote timeout time.  If a cohort crashes before sending a vote to commit, then the blockchain will wait until the vote timeout time and all cohorts will abort the transaction. If a cohort crashes after voting to commit and the blockchain decides that the transaction should commit, then the cohort will replay all transactions that committed while it crashed and recovered. To replay the transactions, the cohort will acquire the same locks it had for the transactions and send the relevant Put and Get requests to the database and commit the transactions in the database. Since it has already computed the response, if the coordinator asks for the get responses for any of these transactions, the cohort can safely return them, even if they have not yet committed in the database. This can be implemented in the future, but we decided not to include it in our first version.

\subsubsection{Slow blockchain}
We use Ethereum for our blockchain implementation. According to [this article], it takes around 15 seconds to 5 minutes to process a blockchain transaction on Ethereum, which indicates that the blockchain is relatively slow compared to other components within our system. A slow blockchain affects the number of commit votes to collect before the vote timeout time, which leads to more transactions to abort eventually; however, the blockchain slowness doesn’t affect the promised atomicity. The client or the coordinator can specify a longer timeout time if needed.

\section{Implementation}
In this section we first describe the implementation of our overall system. Then, we illustrate the implementation of individual components.

\subsection{Environment}
We implemented the coordinator and the cohorts with about 2800 lines of C++ code. For communications between components, we used gPRC and ProtoBuf. For providing the extensibility of swapping different database and blockchain implementations, we provided generic interfaces. We tested end-to-end behavior on a single machine with different processes simulating each component.

\subsubsection{Blockchain}
We wrote our two-phase-commit smart contract with Solidity and developed it with Truffle [\url{https://trufflesuite.com/}]. Truffle provides the development environment for us to compile smart contracts, test it, and deploy it locally for debugging. Once we’ve deployed our smart contract, the coordinator and the cohorts can interact with the contract through a smart contract client library. However, we didn’t find a proper C++ smart contract client library to use, so we use a JavaScript library (web.js[\url{https://web3js.readthedocs.io/en/v1.7.3/}]) instead and add an additional gRPC layer to translate to C++.

While testing in the Truffle development, we learned the gas used for successfully committing a transaction with two cohorts is around 0.33 USD [\href{https://automatedwebtools.com/gwei-to-usd-calculator/}{reference}] (see \ref{appendix} for breakdowns).

\subsubsection{Database}
We designed the DatabaseAdapter interface and implemented the methods with LMDB primitives. LMDB provides process-oriented functions like \texttt{Open()}, \texttt{Begin()}, \texttt{Commit()}, \texttt{Put()}, \texttt{Get()}. Our object-oriented C++ implementation has almost 500 lines of code. We chose LMDB because it is fast, lightweight and easy to test in memory; however, the interface is also extensible to other databases like SQLite [6].

@@

\subsubsection{Cohort}
One optimization we used for the cohort is that the \texttt{PrepareTransactionRequest} includes an indicator if this is the only cohort for the transaction. In such cases, we avoid the expensive requests to the blockchain and the cohort can immediately commit a valid transaction as soon as it finishes preparing it in the database. This is safe because it knows that no other cohort can abort the transaction and that it is guaranteed to be atomic within its own database.

@@

\subsubsection{Latency tradeoffs}
Our approach enables the client to tradeoff between latency and the likelihood of committing a transaction. When the client sets a longer time for timeout, it’s more likely to collect sufficient votes from all cohorts to make a COMMIT decision but may have higher latency if a cohort crashes. There is a limit to the minimum timeout, which can be no shorter than the time it takes to finalize a new blockchain block.

As part of a future effort, we can extend the system to tradeoff between further latency reductions and safety by having cohorts send their votes to the coordinator as well as to the blockchain. If the coordinator receives commit votes from all cohorts, it can update the decision in the blockchain and tell all cohorts to commit. In this procedure, we would not need to wait for a new block to be mined in the blockchain to commit a transaction. The loss of safety comes from edge cases where the blockchain is slow to process the transaction and it ends up not receiving the votes in time and thus deciding to abort. If some cohorts received commit messages from the coordinator and others did not and then asked the blockchain for the decision, the cohorts who received commit messages would commit, while the others would abort the transaction. As we expect this to only affect a very small number of transactions, clients would be able to indicate if their transaction can tolerate a small chance of non-atomicity for much faster latency.

>>>>

\documentclass[11pt,conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[htt]{hyphenat}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

@@

We describe the details of each component in the remaining \ref{components}, the complete flow of handling a successful transaction in \ref{txnflow}, and how we handle failures in \ref{failurecases}.

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{./diagram.pdf}
  \caption{This diagram displays the main components within our architecture and example steps of successfully committing a transaction.}
  \label{fig:Panel}
\end{figure*}

@@

\subsubsection{Coordinator} \label{coordinator}
The coordinator exposes two APIs to the clients:
\begin{itemize}
  \item \texttt{CommitAtomicTransaction( CommitAtomicTransactionRequest, CommitAtomicTransactionResponse)}
  \item \texttt{GetTransactionResult( GetTransactionResultRequest, GetTransactionResultResponse)}
\end{itemize}

@@

\subsubsection{Cohort} \label{cohort}
The cohort exposes two APIs to the coordinator:
\begin{itemize}
  \item \texttt{PrepareTransaction(\\PrepareTransactionRequest, PrepareTransactionResponse)}
  \item \texttt{GetTransactionResult(\\GetTransactionResultRequest, GetTransactionResultResponse)}
\end{itemize}

The \texttt{PrepareTransactionRequest()} contains a \texttt{transaction\_id}, a list of operations of the sub-transaction and the vote timeout time, after which the blockchain will not accept votes. When a cohort receives a \texttt{PrepareTransaction()} request from the coordinator, it takes these steps: 
\begin{enumerate}
  \item Start a new thread in the thread pool.
  \item Create a \texttt{DatabaseAdapter} object.
  \item Start a transaction with \texttt{DatabaseAdapter:: Begin()}.
  \item Acquire the relevant locks for the sub-transaction.
  \item Execute a series of requested \texttt{Put} or \texttt{Get} operations in the sub-transaction. 
  \item If any operation is reported invalid by the database or locks cannot be acquired in time, the cohort will send an \texttt{ABORT} ballot to the blockchain via \texttt{BlockchainAdapter::Vote()}.
  \item If all operations succeed, it will store the retrieved Get values in an in-memory hashmap as well as persistently on disk keyed by the \texttt{transaction\_id} to prepare for future \texttt{GetTransactionResult()} calls.
  \item Set a timer based on the input vote timeout time.
  \item If the timer reaches the vote timeout time, it will send a request \texttt{BlockchainAdapter::\\GetVotingDecision(transaction\_id)} to the blockchain. 
    \begin{enumerate}
      \item If the blockchain states that the transaction should be committed, it will call \texttt{DatabaseAdapter::Commit()} and release any acquired locks.
  \item If the blockchain states that the transaction should be aborted, it will call \texttt{DatabaseAdapter::Abort()} and release any acquired locks.
  \end{enumerate}
\end{enumerate}

When the cohort receives a \texttt{GetTransactionResult()} request and has committed the transaction, it simply reads the in-memory hashmap based on the \texttt{transaction\_id} in the request and returns the got values as response. If it has not yet received a commit or abort response from the blockchain, it returns a pending response instead.

\subsection{System Interactions (a.k.a Life of a Transaction)} \label{txnflow}

This section describes the end-to-end life of a transaction without handling failures.

\begin{enumerate}
  \item The client composes a transaction request and passes it to the coordinator through \texttt{CommitAomticTransaction()}. The coordinator returns a transaction ID for the client to look up the transaction result later.
  \item The coordinator asks the blockchain to trace votes for the new transaction through \texttt{StartVoting()}.
  \item The coordinator decomposes the transaction into sub-transactions and distributes them to the corresponding cohort through \texttt{PrepareTransaction()}.
  \item Each cohort executes a sub-transaction by acquiring locks and forwarding the operations to the database up to the point before Commit(). The cohort writes an entry to the redo log to recover in case it crashes. Additionally, the cohort writes the \texttt{GET} operation results to disk for faster responses after a crash and recovery.
  \item The cohort votes to the blockchain through \texttt{Vote()} indicating that it wants to commit or abort the transaction.
  \item The cohort waits until the expected transaction vote timeout time and gets the voting decision from the blockchain through \texttt{GetVotingDecision()}.
  \item If the voting decision is to commit the transaction, the cohort commits the sub-transaction onto the database and releases any acquired locks; otherwise, it aborts the sub-transaction in the database and releases any acquired locks.
  \item (Independent of the previous steps) The client waits until the vote timeout time and requests the transaction result from the coordinator through \texttt{GetTransactionResult()}, using the transaction ID given in step 1.
  \item The coordinator gets (1) the voting decision of the inquiry transaction from the blockchain through \texttt{GetTransactionResult()} and (2) the \texttt{GET} operation results from each cohort.
\end{enumerate}

@@

\subsubsection{Crash of the coordinator}
The state of the coordinator includes only the list of which cohorts are involved in each active transaction. Thus if the coordinator crashes, when it recovers, it will respond to client requests for the \texttt{Get} responses of the transaction by indicating that it is unable to find the transaction. As a result, the client can directly ask the cohorts what the \texttt{Get} responses of the transaction are. Unlike in traditional 2PC, where the coordinator stores how each cohort has voted, in our design, this information is stored in the blockchain. Our system is resilient to such crashes, as the transaction does not need to block on the coordinator’s recovery, and the cohorts can commit or abort even while the coordinator is unresponsive.

\subsubsection{Crash of a cohort}
Our system is also resilient to crashes of cohorts, as these crashes do not cause any other cohorts to block for longer than the vote timeout time.  If a cohort crashes before sending a vote to commit, then the blockchain will wait until the vote timeout time and all cohorts will abort the transaction. If a cohort crashes after voting to commit and the blockchain decides that the transaction should commit, then the cohort will replay all transactions that committed while it crashed and recovered. To replay the transactions, the cohort will acquire the same locks it had for the transactions and send the relevant \texttt{Put} and \texttt{Get} requests to the database and commit the transactions in the database. Since it has already computed the response, if the coordinator asks for the \texttt{Get} responses for any of these transactions, the cohort can safely return them, even if they have not yet committed in the database. This can be implemented in the future, but we decided not to include it in our first version.

\subsubsection{Slow blockchain}
We use Ethereum for our blockchain implementation. According to \cite{}, it takes around 15 seconds to 5 minutes to process a blockchain transaction on Ethereum, which indicates that the blockchain is relatively slow compared to other components within our system. A slow blockchain increases the likelihood that not all commit votes can be collected before the vote timeout time, which leads to more transactions to abort eventually; however, the blockchain slowness doesn’t affect the promised atomicity. The client or the coordinator can specify a longer vote timeout time if needed.

\section{Implementation}
In this section we first describe the implementation of our overall system. Then, we illustrate the implementation of individual components. The source code of our implementation is available at \cite{} licensed under LGPL. Our implementation depends on these libraries: \cite{}, ...

\subsection{Environment}
We implemented the coordinator and the cohorts with about 2800 lines of C++ code. For communications between components, we used gRPC \cite{} and ProtoBuf \cite{}. For providing the extensibility of swapping different database and blockchain implementations, we provided generic interfaces. We tested end-to-end behaviors on a single machine with different processes simulating each component.

\subsubsection{Blockchain}
We wrote our two-phase-commit smart contract with Solidity and developed it with Truffle \cite{}. Truffle provided the development environment for us to compile smart contracts, test it, and deploy it locally for debugging. Once we deployed our smart contract, the coordinator and the cohorts interacted with the contract through a smart contract client library. However, we didn’t find a proper C++ smart contract client library to use, so we used a JavaScript library (web.js[ref]) instead and added an additional gRPC layer to translate to C++.

While testing in the Truffle development, we learned the gas used for successfully committing a transaction with two cohorts is around 0.33 USD \cite{} (see \ref{appendix} for breakdowns).

\subsubsection{Database}
We designed the \texttt{DatabaseAdapter} interface and implemented the methods with LMDB primitives. LMDB provides process-oriented functions like \texttt{open()}, \texttt{begin()}, \texttt{commit()}, \texttt{put()}, \texttt{get()}. Our object-oriented C++ implementation has almost 500 lines of code. We chose LMDB because it is fast, lightweight and easy to test in memory. The interface is extensible to other databases like SQLite \cite{}.

@@

\subsubsection{Cohort}
One optimization we used for the cohort is that the \texttt{PrepareTransactionRequest()} includes an indicator if this is the only cohort for the transaction. In such cases, we avoid the expensive requests to the blockchain and the cohort can immediately commit a valid transaction as soon as it finishes preparing it in the database. This is safe because it knows that no other cohort can abort the transaction and that it is guaranteed to be atomic within its own database.

@@

\subsubsection{Latency tradeoffs}
Our approach enables the client to tradeoff between latency and the likelihood of committing a transaction. When the client sets a longer time for the vote timeout time, it’s more likely to collect sufficient votes from all cohorts to make a commit decision but may have higher latency if a cohort crashes. There is a limit to the minimum timeout, which can be no shorter than the time it takes to finalize a new blockchain block.

As part of a future effort, we can extend the system to tradeoff between further latency reductions and safety by having cohorts send their votes to the coordinator and to the blockchain. If the coordinator receives commit votes from all cohorts, it can update the decision in the blockchain and tell all cohorts to commit. In this procedure, we would not need to wait for a new block to be mined in the blockchain to commit a transaction. The loss of safety comes from edge cases where the blockchain is slow to process the transaction and it ends up not receiving the votes in time and thus deciding to abort. If some cohorts received commit messages from the coordinator and others did not and then asked the blockchain for the decision, the cohorts who received commit messages would commit, while the others would abort the transaction. As we expect this to only affect a very small number of transactions, clients would be able to indicate if their transaction can tolerate a small chance of non-atomicity for much faster latency.
